---
title: "R Notebook"
output: html_notebook
---

Title : CMS classification for the In-house Dataset using CMScaller i.e. an R package for consensus molecular subtyping of colorectal cancer pre-clinical models. The package has been downloaded from the Lothelab in Github.

1. INSTALLATION OF THE PACKAGES NEEDED FOR THE CMS CLASSIFICATION. 

```{r}
library(readxl)
library(dplyr)
library(org.Hs.eg.db)
library(openxlsx)

#installing all the packages 
new_lib_path <- "C:/Users/Vaishnavi Raikar/Documents/R/win-library/4.3-project"
dir.create(new_lib_path, showWarnings = FALSE, recursive = TRUE)

#setting the custom library path
.libPaths(new_lib_path)

#installing BiocManager and remotes packages.
install.packages("BiocManager", lib = new_lib_path)
install.packages("remotes", lib = new_lib_path)

#installing the CMSclassifier from GITHUB 
remotes::install_github("Sage-Bionetworks/CMSclassifier", lib = new_lib_path)

library(BiocManager, lib.loc = new_lib_path)
library(CMSclassifier, lib.loc = new_lib_path)

#installing the devtools package.
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

#installinhg CMScaller package from GitHub
devtools::install_github("Lothelab/CMScaller")

#loading the CMScaller library
library(CMScaller)

#installing the biomaRt package which will be needed further for the gene identifer conversion.
BiocManager::install("biomaRt")

#loading the biomart package as well.
library(biomaRt)

#loading the necessary packages for loading excel sheet and write the excel sheet as well.
if (!requireNamespace("writexl", quietly = TRUE)) install.packages("writexl")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
install.packages("ggplot2")

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

2. CONVERTING THE GENE SYMBOLS TO ENTREZ IDS. 

```{r}
#loading the in-house dataset
 file_path = "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset.xlsx"
data <- read_excel(file_path)

#extracting the gene symbols from the gene symbol column
gene_symbols <- unique(data$Gene)

#mapping the gene symbols to Entrez IDs using org.Hs.eg.db
mapped_ids <- AnnotationDbi::select(org.Hs.eg.db,
                                    keys = gene_symbols,
                                    columns = c("SYMBOL", "ENTREZID"),
                                    keytype = "SYMBOL")

#handling multiple mappings by keeping the first occurrence for each SYMBOL
mapped_ids_unique <- mapped_ids %>%
  distinct(SYMBOL, .keep_all = TRUE)

#removing the rows with NA Entrez IDs
mapped_ids_clean <- mapped_ids_unique %>%
  filter(!is.na(ENTREZID))

#merging the Entrez IDs back with the original data
merged_data <- left_join(data, mapped_ids_clean, by = c("Gene" = "SYMBOL"))

#removing the rows with NA Entrez IDs from the merged data
final_data <- merged_data %>%
  filter(!is.na(ENTREZID))

output_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset_converted.xlsx"
write.xlsx(final_data, output_path) 

```

3. SETTING THE ENTREZ ID COLUMN AS ROW NAMES FOR THE CMS CLASSIFICATION.

```{r}

#converting the data to a tibble
final_data_1 <- as_tibble(final_data)

#moving the Entrez ID column to be the first column and removing the original Gene column
final_data_1 <- final_data_1 %>%
  dplyr::select(ENTREZID, everything(), -Gene)

#converting the tibble to data frame to set row names
final_data_1 <- as.data.frame(final_data_1)

#finally setting the Entrez ID as row names
rownames(final_data_1) <- final_data$ENTREZID
final_data_1 <- final_data_1 %>% dplyr::select(-ENTREZID)


```


4. PERFORMING THE CMS CLASSIFICATION ON THE IN-HOUSE DATASET.

```{r}

#using the in house data again
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset_converted.xlsx"
final_data_1 <- read_xlsx(file_path)

#converting to the dataframe
final_data_1 <- as.data.frame(final_data_1)

#removing the first column which is the gene column
final_data_1 <- final_data_1[, -1]

#setting the Entrez ID as row names and remove the Entrez ID column from the data
rownames(final_data_1) <- final_data_1$ENTREZID
final_data_1 <- final_data_1[, !(names(final_data_1) %in% "ENTREZID")]

#converting the data into an expression matrix
expression_matrix <- as.matrix(final_data_1)

#performing the CMS classification using CMScaller
cms_results <- CMScaller(expression_matrix, RNAseq = TRUE, doPlot = TRUE)

#lastly extracting the CMS results. 
cms_predictions <- cms_results$prediction


```

5. PLOTTING THE BAR PLOT FOR EACH PATIENT WITH THE COUNT OF SAMPLES BEING CLASSIFIED AS ANY OF THE FOUR CMS CATEGORIES.

```{r}

#using the CMS predictions and storing the counts for patient classification subtype in an excel sheet and then plotting the bar plot.
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/bar_plot.xlsx"
cms_results <- read_xlsx(file_path)

#reshaping the data for plotting
cms_long <- cms_results %>%
  pivot_longer(cols = c(starts_with("CMS"), "NA"), names_to = "CMS", values_to = "Count")

#defining the colors for each CMS group
cms_colors <- c("CMS1" = "#1f77b4", "CMS2" = "#ff7f0e", "CMS3" = "#2ca02c", "CMS4" = "#d62728", "NA" = "#808080")

#creating the bar plot.
ggplot(cms_long, aes(x = Sample_Names, y = Count, fill = CMS)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = cms_colors) +
  labs(title = "CMS Counts per Patient Group",
       x = "Patient Group",
       y = "Count",
       fill = "CMS") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

6. PLOTTING THE DOMINANT SUBTYPE FOR EACH PATIENT. 

```{r}
#defining the corrected dominant subtype data based on the observation meaning if a patient has high number of sample counts for CMS1  , then the dominant subtype will be CMS1.

dominant_subtype <- data.frame(
  Patient = c("Pat_1", "Pat_10", "Pat_2", "Pat_3", "Pat_4", "Pat_5", "Pat_6", "Pat_7", "Pat_8", "Pat_9"),
  Dominant_CMS = c("CMS3", "CMS4", "CMS1", "CMS1", "CMS4", "CMS1", "CMS2", "CMS3", "None", "CMS4"),
  Count = c(3, 4, 5, 5, 4, 2, 3, 5, 0, 4)  #counts per patient sample, so 3 means that there are 3 samples for Pat_1 which has CMS3 classification.
)

#creating the scatter plot with marker size based on the number of samples with dominant subtype
ggplot(dominant_subtype, aes(x = Patient, y = Dominant_CMS, size = Count, color = Dominant_CMS)) +
  geom_point() +
  scale_size_continuous(range = c(3, 10)) +  # Adjust the size range of markers
  labs(title = "Dominant CMS per Patient Group",
       x = "Patient Group",
       y = "Dominant CMS",
       size = "Sample Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

