---
title: "R Notebook"
output: html_notebook
---

Title : CRIS Classification for the In house Dataset. The classification was performed using the predictCRISclassKTSP function which categorizes the expression data into 5 different subtypes. CRISA, CRISB, CRISC, CRISD, and CRISE. 

1. INSTALLING THE PACKAGES.

```{r}

library(readxl)
library(dplyr)
library(org.Hs.eg.db)
library(openxlsx)

#installing all the packages 
new_lib_path <- "C:/Users/Vaishnavi Raikar/Documents/R/win-library/4.3-project"
dir.create(new_lib_path, showWarnings = FALSE, recursive = TRUE)

#setting the custom library path
.libPaths(new_lib_path)

install.packages("BiocManager", lib = new_lib_path)
install.packages("remotes", lib = new_lib_path)

#installing the CMSclassifier from GITHUB 
remotes::install_github("Sage-Bionetworks/CMSclassifier", lib = new_lib_path)

library(BiocManager, lib.loc = new_lib_path)
library(CMSclassifier, lib.loc = new_lib_path)

#devtoolsinstallation
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

#installing CMScaller package from GitHub
devtools::install_github("Lothelab/CMScaller")

# Load necessary libraries
library(CMScaller)

#installing of other necessary packages and the CRISclassifier package. 
if (!requireNamespace("writexl", quietly = TRUE)) install.packages("writexl")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")

library(readxl)
library(writexl)
library(dplyr)
library(CRISclassifier)

```

2. PERFORMING THE CRIS CLASSIFICATION ON THE DATASET AS HERE WE DO NOT NEED CONVERSION OF THE GENE SYMBOLS TO ENTREZ IDS.

```{r}


file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset.xlsx"

final_data <- read_excel(file_path)

#converting the data into a tibble
final_data_1 <- as_tibble(final_data)

#then to a dataframe to set row names
final_data_1 <- as.data.frame(final_data_1)

#setting the Entrez ID as row names
rownames(final_data_1) <- final_data$Gene
final_data_1 <- final_data_1 %>% dplyr::select(-Gene)

#matrix conversion
expression_matrix_1 <- as.matrix(final_data_1)

#performing the CRIS classification using predictCRISclassKTSP function
cris_results <- predictCRISclassKTSP(expression_matrix_1)

print(cris_results)

```

3. PLOTTING THE BAR PLOT FOR EACH PATIENT WITH THE COUNT OF SAMPLES BEING CLASSIFIED AS ANY OF THE FIVE CRIS CATEGORIES.

```{r}

if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
library(readxl)

install.packages("ggplot2")
library("ggplot2")

if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
library(tidyr)

#uploading the cris_plot excel sheet where I stored the counts for each patient.
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/cris_plot.xlsx"

cris_results <- read_xlsx(file_path)

#reshaping the data for plotting
cris_long <- cris_results %>%
  pivot_longer(cols = c(starts_with("CRIS"), "NA"), names_to = "CRIS", values_to = "Count")

#defining the colors for each CRIS group
cris_colors <- c("CRISA" = "#1f77b4", "CRISB" = "#ff7f0e", "CRISC" = "#2ca02c", "CRISD" = "#d62728", "CRISE" = "pink", "NA" = "grey")

#creating the bar plot
ggplot(cris_long, aes(x = Sample_Names, y = Count, fill = CRIS)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = cris_colors) +
  labs(title = "CRIS Counts per Patient Group",
       x = "Patient Group",
       y = "Count",
       fill = "CMS") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

