---
title: "R Notebook"
output: html_notebook
---

Title : CRISS CLASSIFICATION FOR THE EPICC DATA

1. Converting only the CRIS genes to Ensemble ids first. 

```{r}
#instaling biomart 
if (!requireNamespace("biomaRt", quietly = TRUE)) {
  install.packages("biomaRt")
}
library(biomaRt)

#Ensembl BioMart dataset for human genes
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# List of CRIS gene symbols provided by you
cris_genes <- c("ACSM3", "AGR3", "AGT", "AHNAK2", "AKR1C3", "ALDH4A1", "AMT", "ARHGAP23", "ASCL2", "ATOH1", 
                "BCAS1", "CAB39L", "CALB2", "CDHR1", "CDR2L", "CEACAM7", "CES1", "CKMT2", "CLIC3", "COL17A1", 
                "CRYL1", "CTSA", "CYP4X1", "DPEP1", "ERP27", "FAM69B", "FCGBP", "FGGY", "FYN", "GCNT3", "GGT6", 
                "GUCY2C", "HEPACAM2", "IFIT3", "IL1R2", "IRF9", "KLK1", "KLK10", "KLK6", "KRT23", "LAMA5", 
                "LARP6", "MALL", "MAP7D2", "MLPH", "MRAP2", "MUC12", "MUC2", "MUC4", "MX2", "NES", "NKD1", 
                "NMNAT3", "NOTUM", "PDZK1IP1", "PLTP", "PPP2R2C", "PRSS23", "PTK7", "PTRF", "RAB27B", "REG4", 
                "S100A4", "SERPINB5", "SLC26A3", "SLC3A1", "SLC46A3", "SLC5A1", "SPINK4", "ST6GAL1", "TCN1", 
                "TESC", "TMEM176A", "TMPRSS3", "TNNC2", "TRIM7", "TRNP1", "TSPAN33", "UCA1", "VANGL2")

# Query Ensembl to map the CRIS gene symbols to their corresponding Ensembl IDs
gene_mapping <- getBM(attributes = c("hgnc_symbol", "ensembl_gene_id"), 
                      filters = "hgnc_symbol", 
                      values = cris_genes, 
                      mart = ensembl)

# Print the mapping of CRIS gene symbols to Ensembl IDs
print(gene_mapping)



```
2. Now getting the expression data for these genes and saving it in a new dataframe. 

```{r}

#EPICC dataset
epicc_data <- read_excel("C:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/EPICC_dataset.xlsx")

#filtering the EPICC dataset to retain only rows that match the Ensembl IDs of the CRIS genes
filtered_data <- epicc_data[epicc_data$GeneID %in% gene_mapping$ensembl_gene_id, ]

filtered_data$GeneID <- gene_mapping$hgnc_symbol[match(filtered_data$GeneID, gene_mapping$ensembl_gene_id)]

#renaming the gene symbol column 
colnames(filtered_data)[1] <- "GENE_SYMBOL"

```

3. Only performing the CRIS classification on the cris data. 


```{r}
#loading the classifier 
library(CRISclassifier)

#converting the filtered_data to a data.frame
filtered_data_2 <- as.data.frame(filtered_data)

#setting the Gene Symbol as rownames and removing the first column whihc is the gene symbol column 
rownames(filtered_data_2) <- filtered_data_2$`GENE_SYMBOL`
filtered_data_2 <- filtered_data_2[, -1]  

#converting the filtered data into a matrix for CRIS classification
gene_expression_matrix <- as.matrix(filtered_data_2)

print(dim(gene_expression_matrix))
print(head(rownames(gene_expression_matrix))) 

#performing the CRIS classification
cris_classification <- predictCRISclassKTSP(gene_expression_matrix)

print(cris_classification)

```

4. Plotting them in a bar plolt - only cris genes 

```{r}
#convert the CRIS classification results into a dataframe
cris_classification_df <- data.frame(
  Sample = sub("_.*", "", names(cris_classification$tspSetClassPredsFinal)),  # Extract patient names (before the '_')
  CRIS = as.character(cris_classification$tspSetClassPredsFinal)  # Convert CRIS classifications to characters
)

#group by Sample and CRIS type, and count the occurrences
cris_summary <- cris_classification_df %>%
  group_by(Sample, CRIS) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  tidyr::spread(CRIS, Count, fill = 0)

#ensure all CRIS categories are included
expected_cris <- c("CRISA", "CRISB", "CRISC", "CRISD", "CRISE")
for (cris in expected_cris) {
  if (!cris %in% colnames(cris_summary)) {
    cris_summary[[cris]] <- 0  # Fill missing CRIS categories with zero counts
  }
}

#reordering the columns to display Sample first, followed by CRISA to CRISE
cris_summary <- cris_summary %>%
  dplyr::select(Sample, all_of(expected_cris))

print(cris_summary)

```

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Reshaping the data: Convert from wide to long format
cris_long <- cris_summary %>%
  pivot_longer(cols = -Sample, names_to = "CRIS_Type", values_to = "Count")

# Creating the bar plot
ggplot(cris_long, aes(x = Sample, y = Count, fill = CRIS_Type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "CRIS classification plot for EPICC data", 
       x = "Sample", 
       y = "Count", 
       fill = "CRIS Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("CRISA" = "blue", "CRISB" = "green", "CRISC" = "red", "CRISD" = "yellow", "CRISE" = "purple"))

```

5. Performing the cris classification on the entire dataset 


```{r}
library(biomaRt)
library(dplyr)
library(readxl)
library(writexl)

# 1. Load the EPICC dataset from the specified file path
file <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/EPICC_dataset.xlsx"
data <- read_xlsx(file)

# 2.defining the  Ensembl IDs to remove from the dataset
remove_ids <- c("ENSG00000005187", "ENSG00000173467", "ENSG00000135744", "ENSG00000185567", 
                "ENSG00000196139", "ENSG00000159423", "ENSG00000145020", "ENSG00000273780", 
                "ENSG00000275832", "ENSG00000183734", "ENSG00000172238", "ENSG00000064787", 
                "ENSG00000102547", "ENSG00000282830", "ENSG00000172137", "ENSG00000148600",
                "ENSG00000109089", "ENSG00000280501", "ENSG00000007306", "ENSG00000262243",
                "ENSG00000198848", "ENSG00000131730", "ENSG00000169583", "ENSG00000065618",
                "ENSG00000165475", "ENSG00000064601", "ENSG00000186377", "ENSG00000015413",
                "ENSG00000139055", "ENSG00000281123", "ENSG00000275395", "ENSG00000172456",
                "ENSG00000010810", "ENSG00000140297", "ENSG00000167741", "ENSG00000070019", 
                "ENSG00000188175", "ENSG00000119917", "ENSG00000115590", "ENSG00000285048",
                "ENSG00000213928", "ENSG00000167748", "ENSG00000129451", "ENSG00000167755",
                "ENSG00000263309", "ENSG00000108244", "ENSG00000130702", "ENSG00000166173",
                "ENSG00000144063", "ENSG00000184368", "ENSG00000115648", "ENSG00000135324", 
                "ENSG00000205277", "ENSG00000278466", "ENSG00000198788", "ENSG00000293532",
                "ENSG00000277585", "ENSG00000273822", "ENSG00000278303", "ENSG00000278468",
                "ENSG00000273984", "ENSG00000276613", "ENSG00000275164", "ENSG00000145113",
                "ENSG00000183486", "ENSG00000132688", "ENSG00000140807", "ENSG00000163864",
                "ENSG00000185269", "ENSG00000162366", "ENSG00000100979", "ENSG00000074211",
                "ENSG00000150687", "ENSG00000112655", "ENSG00000041353", "ENSG00000134193",
                "ENSG00000196154", "ENSG00000206075", "ENSG00000091138", "ENSG00000138079",
                "ENSG00000139508", "ENSG00000100170", "ENSG00000122711", "ENSG00000073849",
                "ENSG00000134827", "ENSG00000088992", "ENSG00000002933", "ENSG00000160183",
                "ENSG00000101470", "ENSG00000146054", "ENSG00000253368", "ENSG00000158457",
                "ENSG00000214049", "ENSG00000162738")

#filter out the Ensembl IDs to be removed
data_filtered <- data %>%
  filter(!GeneID %in% remove_ids)

#connect to Ensembl via biomaRt to get gene symbols for the remaining Ensembl IDs
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

#extract the unique Ensembl IDs from the remaining dataset
ensembl_ids <- unique(data_filtered$GeneID)

#split the Ensembl IDs into batches to avoid server timeouts
batch_size <- 500
ensembl_id_batches <- split(ensembl_ids, ceiling(seq_along(ensembl_ids) / batch_size))

#initialize an empty dataframe to store the mapping results
mapped_ids <- data.frame()

#loop through each batch and query Ensembl for gene symbols
for (batch in ensembl_id_batches) {
  tryCatch({
    batch_result <- getBM(
      attributes = c('ensembl_gene_id', 'hgnc_symbol'),
      filters = 'ensembl_gene_id',
      values = batch,
      mart = ensembl
    )
    
    #append the results to the mapped_ids dataframe
    mapped_ids <- bind_rows(mapped_ids, batch_result)
  
    Sys.sleep(1)
    
  }, error = function(e) {
    message("Error: ", e)
  })
}

#renaming the columns 
colnames(mapped_ids) <- c('GeneID', 'GENE_SYMBOL')

#removing the duplicates and ensure that gene symbols are unique
mapped_ids_unique <- mapped_ids %>%
  group_by(GeneID) %>%
  summarise(GENE_SYMBOL = dplyr::first(GENE_SYMBOL)) %>%
  ungroup()

#removing the rows with NA Gene Symbols
mapped_ids_clean <- mapped_ids_unique %>%
  filter(!is.na(GENE_SYMBOL))

#merging the mapped gene symbols back with the filtered data
merged_data <- left_join(data_filtered, mapped_ids_clean, by = "GeneID")

#defining the expression columns (excluding the first column which is GeneID)
expression_cols <- names(data)[2:ncol(data)]

#averaging the expression values where multiple Ensembl IDs map to the same gene symbol
final_data <- merged_data %>%
  group_by(GENE_SYMBOL) %>%
  summarise(across(all_of(expression_cols), mean, na.rm = TRUE)) %>%
  ungroup()



```


6. MERGING THE TWO DATASETS - ONE WITH CRIS GENES expression data and the other without those genes but the rest of the data because we already did the mapping for cris genes seperately.

```{r}

# Combine final_data and filtered_data based on GENE_SYMBOL
combined_data <- bind_rows(final_data, filtered_data)

str(combined_data)
print(head(combined_data))

output = "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/EPICC_dataset_genesymbols_combined.xlsx"
write.xlsx(combined_data, output)


```
7. Performing cris classification now for this dataset. 

```{r}
#checking if there are any missing values in the 'GENE_SYMBOL' column
sum(is.na(combined_data$GENE_SYMBOL))

#if there are any missing values removing them before setting the genes as row names 
combined_data <- combined_data %>% filter(!is.na(GENE_SYMBOL))
rownames(combined_data) <- combined_data$GENE_SYMBOL

#cris classification
gene_expression_matrix <- as.matrix(combined_data)

# Perform CRIS classification
cris_classification_epicc <- predictCRISclassKTSP(gene_expression_matrix)

# Print the classification results
print(cris_classification_epicc)


```

8. PLOTTING THE GRAPH FOR THE CRISS CLASSIFICATION. 

```{r}

#convert the CRIS classification results into a dataframe
cris_classification_epic<- data.frame(
  Sample = sub("_.*", "", names(cris_classification_epicc$tspSetClassPredsFinal)),  # Extract patient names (before the '_')
  CRIS = as.character(cris_classification_epicc$tspSetClassPredsFinal)  # Convert CRIS classifications to characters
)

#group by Sample and CRIS type, and count the occurrences
cris_summary_epic <- cris_classification_epic %>%
  group_by(Sample, CRIS) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  tidyr::spread(CRIS, Count, fill = 0)

#ensure all CRIS categories (A-E) are included
expected_cris_epic <- c("CRISA", "CRISB", "CRISC", "CRISD", "CRISE")
for (cris in expected_cris_epic) {
  if (!cris %in% colnames(cris_summary_epic)) {
    cris_summary_epic[[cris]] <- 0  # Fill missing CRIS categories with zero counts
  }
}

#reordering the columns to display Sample first, followed by CRISA to CRISE
cris_summary_epic <- cris_summary_epic %>%
  dplyr::select(Sample, all_of(expected_cris_epic))


#removing the row where Sample is "GENE"
cris_summary_epic <- cris_summary_epic %>% 
  filter(Sample != "GENE")
print(cris_summary_epic)

```


```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

#reshaping the data 
#Converting from wide to long format
cris_long <- cris_summary_epic %>%
  pivot_longer(cols = -Sample, names_to = "CRIS_Type", values_to = "Count")

#creating the bar plot
ggplot(cris_long, aes(x = Sample, y = Count, fill = CRIS_Type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "CRIS classification plot for entire EPICC data ", 
       x = "Sample", 
       y = "Count", 
       fill = "CRIS Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("CRISA" = "blue", "CRISB" = "green", "CRISC" = "red", "CRISD" = "yellow", "CRISE" = "purple"))
```

