---
title: "R Notebook"
output: html_notebook
---

1. INSTALLING THE CMS PACKAGE FOR EPICC DATA.  

```{r}

library(readxl)
library(dplyr)
library(org.Hs.eg.db)
library(openxlsx)
library(biomaRt)
library(tidyr)

#installing all the packages 
new_lib_path1 <- "C:/Users/Vaishnavi Raikar/Documents/R/win-library/4.3-project"
dir.create(new_lib_path1, showWarnings = FALSE, recursive = TRUE)

#setting the custom library path
.libPaths(new_lib_path1)

install.packages("BiocManager", lib = new_lib_path1)
install.packages("remotes", lib = new_lib_path1)

#installing the CMSclassifier from GITHUB 
remotes::install_github("Sage-Bionetworks/CMSclassifier", lib = new_lib_path1)

library(BiocManager, lib.loc = new_lib_path1)
library(CMSclassifier, lib.loc = new_lib_path1)

if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

#CMScaller package from GitHub
devtools::install_github("Lothelab/CMScaller")
library(CMScaller)


```

2. INSTALLING THE PACKAGES FOR CONVERTING THE ENSEMBLE IDS TO ENTREZ IDS FOR EPICC DATA.

```{r}

#installing the bioconductor package 
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

#installation of the biomart package 
BiocManager::install("biomaRt")

#loading biomart package 
library(biomaRt)

if (!requireNamespace("writexl", quietly = TRUE)) install.packages("writexl")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)

```

3. MAPPING THE ENSEMBL IDS TO ENTREZ IDS AND HANNDLING DUPLICATE IDS BY MERGING THE EXPRESSION VALUES ( BY TAKING THE AVERAGE ) OF THE ENSEMBLE IDS WHICH MAPS TO THE TWO ENTREZ IDS. 

```{r}
library(readxl)
data <- read_excel("C:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/EPICC_dataset.xlsx")

#connecting to ensembl via biomart
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

#extracting the unique Ensembl IDs from the dataset 
ensembl_ids <- unique(data$GeneID)

#mapping the Ensembl IDs to Entrez IDs using biomaRt
mapped_ids <- getBM(
  attributes = c('ensembl_gene_id', 'entrezgene_id'),
  filters = 'ensembl_gene_id',
  values = ensembl_ids,
  mart = ensembl
)

#renaming the columns for consistency
colnames(mapped_ids) <- c('GeneID', 'ENTREZID')

#ensuring that ENTREZID is a character 
mapped_ids$ENTREZID <- as.character(mapped_ids$ENTREZID)

#removing the duplicates by keeping the first occurrence of each Ensembl ID to Entrez ID mapping
mapped_ids_unique <- mapped_ids %>%
  group_by(GeneID) %>%
  summarise(ENTREZID = dplyr::first(ENTREZID)) %>%
  ungroup()

#removing the rows with NA Entrez IDs
mapped_ids_clean <- mapped_ids_unique %>%
  filter(!is.na(ENTREZID))

#merging the Entrez IDs back with the original data using Ensembl IDs
merged_data <- left_join(data, mapped_ids_clean, by = "GeneID")

#averaging the expression values where multiple Ensembl IDs map to the same Entrez ID
expression_cols <- names(data)[2:ncol(data)]  

final_data <- merged_data %>%
  group_by(ENTREZID) %>%
  summarise(across(all_of(expression_cols), mean, na.rm = TRUE)) %>%
  ungroup()

output_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/EPICC_dataset_converted.xlsx"
write.xlsx(final_data, output_path)

```

4. PERFORMING THE CMS CLASSIFICATION ON THE EPICC DATASET. 

```{r}

gene_ids <- final_data$ENTREZID

#extracting the expression data apart from the first column. 
expr_matrix <- as.matrix(final_data[, -1])

#checking if the dimensions match 
cat("Rows in expr_matrix: ", nrow(expr_matrix), "\n")
cat("Length of gene_ids: ", length(gene_ids), "\n")

#setting the row names of expr_matrix to the Entrez IDs
rownames(expr_matrix) <- gene_ids

#performing the CMS classification
cms_results <- CMScaller(expr_matrix, RNAseq = TRUE, doPlot = TRUE)


print(cms_results)


```
5. MAKING A TABLE WITH THE COUNT OF HOW MANY SAMPLES HAVE THE MOLECULAR SUBTYPES. 

```{r}
library(tidyr)

#extracing the sample names from row_names 
cms_results$Sample <- sub("_.*", "", rownames(cms_results))

#replacing NA prediction with string "NA"
cms_results$prediction <- as.character(cms_results$prediction)
cms_results$prediction[is.na(cms_results$prediction)] <- "NA"

#grouping by the sample and the CMS subtype prediction to count the occurrences
cms_summary <- cms_results %>%
  group_by(Sample, prediction) %>%
  summarise(count = n(), .groups = 'drop') %>%
  tidyr::spread(prediction, count, fill = 0)

#filling missing CMS categories with zeros if they are missing in any sample
expected_cms <- c("CMS1", "CMS2", "CMS3", "CMS4", "NA")
for (cms in expected_cms) {
  if (!cms %in% colnames(cms_summary)) {
    cms_summary[[cms]] <- 0
  }
}

#ungrouping the data before select() function
#using dplyr::select() explicitly
cms_summary <- cms_summary %>%
  ungroup() %>%
  dplyr::select(Sample, all_of(expected_cms)) 

print(cms_summary)

```


6. PLOTTING THE CMS PLOT FOR THE EPICC DATASET 

```{r}
library(ggplot2)
#reshaping the data
cms_long <- cms_summary %>%
  pivot_longer(cols = -Sample, names_to = "CMS_Type", values_to = "Count")

#creating the bar plot
ggplot(cms_long, aes(x = Sample, y = Count, fill = CMS_Type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "CMS classification plot for EPICC data", 
       x = "Sample", 
       y = "Count", 
       fill = "CMS Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("CMS1" = "blue", "CMS2" = "violet", "CMS3" = "red", "CMS4" = "yellow", "NA" = "gray"))
```






