---
title: "R Notebook"
output: html_notebook
---

MEAN EXPRESSION ANALYSIS FOR THE IN-HOUSE DATASET TO UNDERSTAND THE VARIABILITY BETWEEN THE SAMPLES USING COEFFICIENT OF VARIATION METHOD.

-   Pre-processing the data to only include the highly expressed genes.
-   Calculating the Coefficient of Variation ( CV ) for each patient using for loop by using the formula CV = Variance / Mean.
-   Lastly, plotting the average CV across all patients using ggplot.

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(readxl)

#calculating the CV for a given patient group
calculate_cv_for_patient <- function(patient_data) {
  #removing the genes with 0 counts in all patient samples
  filtered_data <- patient_data %>%
    filter(rowSums(dplyr::select(., -Gene) != 0) > 0)
  
  #removing the genes with less than 20 counts in all patient samples
  filtered_data <- filtered_data %>%
    filter(rowSums(dplyr::select(., -Gene) >= 20) > 0)
  
  #calculating  the mean, variance, and CV
  pat_stats <- filtered_data %>%
    rowwise() %>%
    mutate(Mean = mean(c_across(-Gene), na.rm = TRUE),
           Variance = var(c_across(-Gene), na.rm = TRUE),
           CV = sqrt(Variance) / Mean) %>%
    ungroup()
  
  return(pat_stats)
}

#loading the In house dataset 
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset.xlsx"
all_data <- read_xlsx(file_path)

#removing the entrez IDs 
col_names <- colnames(all_data)
col_names <- col_names[!grepl("ENTREZID", col_names)]  #excluding the ENTREZID column

#creating empty lists to store the average CV and Mean for each patient
average_cv_list <- list()
average_mean_list <- list()

#looping through each unique patient identifier
for (pat in unique(gsub("_Sam.*", "", col_names[-1]))) {
  #selecting the columns related to the patient data
  pat_cols <- col_names[grep(paste0("^", pat, "_"), col_names)]
  
  #creating a subste of the data for that specific patient  
  pat_data <- all_data %>%
    dplyr::select(Gene, all_of(pat_cols))
  
  #calculating the statistics for pat_data
  pat_stats <- calculate_cv_for_patient(pat_data)
  
  #the calculating the average CV and mean for this patient and store it
  average_cv <- mean(pat_stats$CV, na.rm = TRUE)
  average_mean <- mean(pat_stats$Mean, na.rm = TRUE)
  average_cv_list[[pat]] <- average_cv
  average_mean_list[[pat]] <- average_mean
}

#converting the list to a data frame for better plotting
average_cv_df <- data.frame(
  Patient = names(average_cv_list),
  Average_CV = unlist(average_cv_list),
  Average_Mean = unlist(average_mean_list)
)

#plotting the average CV for all patients
p1 <- ggplot(average_cv_df, aes(x = Patient, y = Average_CV, fill = Patient)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Average Coefficient of Variation (CV) for Each Patient Group",
       x = "Patient Group",
       y = "Average CV") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")

print(p1)


```

2.  PLOTTING THE SCATTER PLOT FOR MEAN EXPRESSION VS THE CV TO CONSIDER THE CRITERIA TO SELECT THE THRESHOLD.

-   Most of the genes have mean_expression less than 100 , so if we remove those genes , then we will have a linear distribution of the CVs.
-   The threshold was selected as " Mean_expression < 100 ". So if the genes have less than 100 mean expression , they have been discarded.

```{r}

#calculate CV for a given patient group without filtering
calculate_cv_for_patient <- function(all_data1, pat) {
  #removing the genes with 0 counts in all patient samples
  filtered_dataa <- all_data1 %>%
    filter(rowSums(dplyr::select(., -Gene) != 0) > 0)
  
  #reemoving the genes with less than 20 counts in all the samples
  filtered_dataa <- filtered_dataa %>%
    filter(rowSums(dplyr::select(., -Gene) >= 20) > 0)
  
  #calculation of mean, variance, and CV
  pat_statss <- filtered_dataa %>%
    rowwise() %>%
    mutate(Mean = mean(c_across(-Gene), na.rm = TRUE),
           Variance = var(c_across(-Gene), na.rm = TRUE),
           CV = sqrt(Variance) / Mean,
           Patient = pat) %>%  # Add the patient identifier as a new column
    ungroup()
  
  return(pat_statss)
}


file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset.xlsx"
all_data1 <- read_xlsx(file_path)

#fetching the column names 
col_names <- colnames(all_data1)

#defining a color pallete for the graph 
unique_patients <- unique(gsub("_Sam.*", "", col_names[-1]))
color_palette <- scales::brewer_pal(palette = "Set3")(length(unique_patients))

#looping through each unique patient identifier
for (i in seq_along(unique_patients)) {
  pat <- unique_patients[i]
  
  #selecting the columns related to the patient data
  pat_cols <- col_names[grep(paste0("^", pat, "_"), col_names)]
  
  #creating a subset of the data for the specific patient
  pat_dataa <- all_data1 %>%
    dplyr::select(Gene, all_of(pat_cols))
  
  #calculating the CV for the patient without filtering
  cv_data <- calculate_cv_for_patient(pat_dataa, pat)
  
  #lastly generating the scatter plot for each patient group with log scales
  p <- ggplot(cv_data, aes(x = Mean, y = CV)) +
    geom_point(alpha = 0.7, color = color_palette[i]) +
    scale_x_log10() +  # Log scale for x-axis (Mean Expression)
    scale_y_log10() +  # Log scale for y-axis (CV)
    theme_minimal() +
    labs(title = paste("Log-Scale Scatter Plot of Mean Expression vs. CV for", pat),
         x = "Mean Expression (log scale)",
         y = "Coefficient of Variation (CV, log scale)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)  
}

```
