---
title: "R Notebook"
output: html_notebook
---


MEAN EXPRESSION ANALYSIS ON EPICC DATA.


-  First I have pre processed the data to only include the highly expressed genes. The filtering criteria is different for this dataset as the dataset is large and it contains a lot of lowely expressed genes. 
-  Then I have calculated the Coefficient of Variation ( CV ) for each patient using for loop by using the formula CV = Variance / Mean. 
-  Lastly, plotting the average CV across all patients using ggplot. 

Note :  Patient C555 and C562 have  only one sample i.e.C555_C1_G2. The Coefficient of Variation (CV) requires at least two values (samples) to calculate variance. When you have only one value, the variance is undefined, and this leads to NA values for the variance and thus the CV is 0. 


```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(readxl)
library(viridis)

#calculating the Coefficient of Variation for each patient group. 
calculate_cv_for_patient <- function(patient_data) {
  #cv = variance / mean 
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(Mean = mean(c_across(-GeneID), na.rm = TRUE),
           Variance = var(c_across(-GeneID), na.rm = TRUE),
           CV = sqrt(Variance) / Mean) %>%
    ungroup()
  
  return(pat_stats)
}

#loading the EPICC dataset
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/EPICC_dataset.xlsx"
all_data <- read_xlsx(file_path)

#filtering the dataset using these criterias to only keep the highly expressed genes. 
filtered_data <- all_data %>%
  rowwise() %>%
  filter(sum(c_across(starts_with("C"))) > 0) %>%
  filter(sum(c_across(starts_with("C"))) >= 30) %>%
  filter(mean(c_across(starts_with("C")) > 50) >= 0.5)


col_names <- colnames(filtered_data)

#extracting the patient identifier names
#by taking the first part of each column ( C516 from C516_B1_G7)
patient_ids <- unique(sub("_.*", "", col_names[-1]))

#creating empty lists to store the avg CV and mean for each of the patients. 
average_cv_list <- list()
average_mean_list <- list()

#looping through each unique patient identifier
for (pat in patient_ids) {
  #selecting all columns related to the patient. 
  pat_cols <- grep(paste0("^", pat, "_"), colnames(filtered_data), value = TRUE)
  
  #checking if the patient column exists  
  if (length(pat_cols) > 0) {
    #creating a subset for that specific patient 
    pat_data <- filtered_data %>%
      dplyr::select(GeneID, all_of(pat_cols))  
    
    #calculating the stats for that patient 
    pat_stats <- calculate_cv_for_patient(pat_data)
    
    #calculating the average CV and mean for this patient and storing it
    average_cv <- mean(pat_stats$CV, na.rm = TRUE)
    average_mean <- mean(pat_stats$Mean, na.rm = TRUE)
    
    average_cv_list[[pat]] <- average_cv
    average_mean_list[[pat]] <- average_mean
  }
}

#converting the list to a dataframe for plotting
average_cv_df <- data.frame(
  Patient = names(average_cv_list),
  Average_CV = unlist(average_cv_list),
  Average_Mean = unlist(average_mean_list)
)

#check if any patients are missing : debugging here 
missing_patients <- setdiff(patient_ids, names(average_cv_list))
if (length(missing_patients) > 0) {
  print(paste("Missing patients:", paste(missing_patients, collapse = ", ")))
}

#plotting average CV for all patients 
p1 <- ggplot(average_cv_df, aes(x = Patient, y = Average_CV, fill = Patient)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Average Coefficient of Variation (CV) for Each Patient Group",
       x = "Patient Group",
       y = "Average CV") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()

print(p1)


```

2. Plotting the scatter plot for each patient group with log scales. Mean Expression vs CV. 

```{r}

library(dplyr)
library(ggplot2)
library(readr)
library(readxl)
library(viridis)

#calculating the coefficient of variation (CV) for each patient group
calculate_cv_for_patient <- function(patient_data) {
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(Mean = mean(c_across(-GeneID), na.rm = TRUE),
           Variance = var(c_across(-GeneID), na.rm = TRUE),
           CV = sqrt(Variance) / Mean) %>%
    ungroup()
  return(pat_stats)
}

file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/EPICC_dataset.xlsx"  # Adjusted for the uploaded file
all_data <- read_xlsx(file_path)

# Filter the dataset using the specified criteria to only keep the highly expressed genes
filtered_data <- all_data %>%
  rowwise() %>%
  filter(sum(c_across(starts_with("C"))) > 0) %>%
  filter(sum(c_across(starts_with("C"))) >= 30) %>%
  filter(mean(c_across(starts_with("C")) > 50) >= 0.5)

col_names <- colnames(filtered_data)

#extracting patient identifier names by taking the first part of each column name (e.g., C516 from C516_B1_G7)
patient_ids <- unique(sub("_.*", "", col_names[-1]))

#creating empty lists to store the avg CV and mean for each of the patients
average_cv_list <- list()
average_mean_list <- list()

#defining the color pallete 
color_palette <- viridis::viridis(length(patient_ids))

#looping through each unique patient identifier
for (i in seq_along(patient_ids)) {
  pat <- patient_ids[i]
  
  pat_cols <- grep(paste0("^", pat, "_"), colnames(filtered_data), value = TRUE)
  
  #checking if the patient column exists
  if (length(pat_cols) > 0) {
    #creating a subset for that specific patient
    pat_data <- filtered_data %>%
      dplyr::select(GeneID, all_of(pat_cols))
    
    #calculating the stats for that patient
    pat_stats <- calculate_cv_for_patient(pat_data)
    
    #filtering out rows with Mean or CV values near zero to avoid log issues
    pat_stats <- pat_stats %>%
      filter(Mean > 500 , CV > 0.2 )  

    #plotting the scatter plot for this patient with a unique color
    p <- ggplot(pat_stats, aes(x = log(Mean), y = log(CV))) +
      geom_point(alpha = 0.5, color = color_palette[i]) +  # Assign a unique color
      theme_minimal() +
      labs(title = paste("Log-Scale Scatter Plot of Mean Expression vs. CV for", pat),
           x = "Mean Expression (log scale)",
           y = "Coefficient of Variation (CV, log scale)") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    print(p)
  }
}


```
```{r}

library(dplyr)

#calculating mean, variance, and CV for each patient group
calculate_cv_for_patient <- function(patient_data) {
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(Mean = mean(c_across(-GeneID), na.rm = TRUE),
           Variance = var(c_across(-GeneID), na.rm = TRUE),
           CV = sqrt(Variance) / Mean) %>%
    ungroup()
  return(pat_stats)
}

#looping through each unique patient identifier and print the dataframe
for (pat in patient_ids) {
  
  #selecting all columns related to the patient
  pat_cols <- grep(paste0("^", pat, "_"), colnames(filtered_data), value = TRUE)
  
  #checking if the patient column exists
  if (length(pat_cols) > 0) {
    pat_data <- filtered_data %>%
      dplyr::select(GeneID, all_of(pat_cols))
    pat_stats <- calculate_cv_for_patient(pat_data)
    cat("\nData for patient:", pat, "\n")
    print(pat_stats)
  }
}

```



3. PERFORMING LOOCV FOR PATIENT 1 - 10 


```{r}
library(dplyr)
library(readxl)
library(ggplot2)
library(viridis)

file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/EPICC_dataset.xlsx"
all_data <- read_xlsx(file_path)

#filtering the dataset to keep only the highly expressed genes
filtered_data <- all_data %>%
  rowwise() %>%
  filter(sum(c_across(starts_with("C"))) > 0) %>%
  filter(sum(c_across(starts_with("C"))) >= 30) %>%
  filter(mean(c_across(starts_with("C")) > 50) >= 0.5)

#extracting unique patient identifiers by taking the first part of the column names
patient_groups <- unique(gsub("_.*", "", colnames(filtered_data)[-1]))

#limiting the analysis first to the first 10 patient IDs
patient_groups <- patient_groups[1:10]

#initializing a list to store the LOOCV data frames
loocv_dfs <- list()
average_cv_dfs <- data.frame()

# Function to calculate the mean, variance, and CV for a given patient's data
calculate_stats <- function(patient_data) {
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(
      Mean = mean(c_across(where(is.numeric)), na.rm = TRUE),
      Variance = var(c_across(where(is.numeric)), na.rm = TRUE),
      CV = ifelse(Mean == 0, NA, sqrt(Variance) / Mean)  # Avoid division by zero
    ) %>%
    ungroup()
  
  return(pat_stats)
}

# Loop through the first 10 patient groups to perform LOOCV
for (pat in patient_groups) {
  # Match the samples for each patient
  pat_data_filtered <- filtered_data %>%
    dplyr::select(GeneID, matches(paste0("^", pat, "_")))

  # Get the column names corresponding to the patient samples
  pat_columns <- colnames(pat_data_filtered)[grep(paste0("^", pat, "_"), colnames(pat_data_filtered))]
  
  # Calculating the row-wise means (mean expression) across patient columns
  row_means <- rowMeans(pat_data_filtered[, -1], na.rm = TRUE)
  
  # Calculating the median of the row-wise means
  median_threshold <- median(row_means, na.rm = TRUE)
  
  # Initializing a list to store CV values for each LOOCV iteration
  cv_values <- list()
  
  # Perform LOOCV for each sample in the patient group
  for (i in seq_along(pat_columns)) {
    # Exclude one sample at a time and calculate stats for the remaining samples
    loocv_data <- pat_data_filtered %>% dplyr::select(-all_of(pat_columns[i]))
    
    #calculate Mean, Variance, and CV for the remaining samples
    stats <- calculate_stats(loocv_data)
    
    # Filter out genes with mean expression less than the median threshold
    stats_filtered <- stats %>%
      filter(Mean >= median_threshold)
    
    #only the top 500 genes with the highest CV
    top_stats <- stats_filtered %>%
      arrange(desc(CV)) %>%
      slice_head(n = 500)
    if (nrow(top_stats) > 0) {  
      cv_column_name <- paste(pat_columns[i], "CV", sep = "_")
      loocv_dfs[[cv_column_name]] <- top_stats$CV
      cv_values[[cv_column_name]] <- top_stats$CV
    }
  }
  
  #combine the CV values into a data frame if there are CV values
  if (length(cv_values) > 0) {
    cv_df <- as.data.frame(cv_values)
    
    # Calculate the average CV for each LOOCV iteration
    avg_cv <- colMeans(cv_df, na.rm = TRUE)
    
    # Ensure avg_cv is not empty before proceeding
    if (length(avg_cv) > 0) {
      # Combine the average CV into the result data frame
      avg_cv_df <- data.frame(Patient = pat, Sample = names(avg_cv), Avg_CV = avg_cv)
      
      # Append the results to the main average CV dataframe
      average_cv_dfs <- rbind(average_cv_dfs, avg_cv_df)
    }
  }
  
  #print the resulting data frames for each patient
  print(paste("Data frame for patient:", pat))
  print(cv_df)
}

library(ggplot2)

#creating a dynamic color scale based on the number of unique samples
unique_samples <- unique(average_cv_dfs$Sample)
color_count <- length(unique_samples)
color_palette <- scales::hue_pal()(color_count)

ggplot(average_cv_dfs, aes(x = Patient, y = Avg_CV, color = Sample)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Average CV per Patient",
       x = "Patient Group",
       y = "Average CV") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = color_palette)  # Apply dynamic colors

#saving the plot
ggsave("C:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/Average_CV_per_Patient_1st_batch.png", width = 15, height = 8, units = "in", dpi = 300, bg= "white")



```
