---
title: "R Notebook"
output: html_notebook
---

Title : Gene Ontology Analysis for the EPICC dataset. 

1. Removing the samples outliers for the first set of LOOCV code and printing those dataframes. 

```{r}
library(dplyr)
library(readxl)
library(ggplot2)
library(viridis)

file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/EPICC_genesymbols_cleaned.xlsx" 
all_data_gene <- read_xlsx(file_path)

#making the the last column the first, and drop the original first column
all_data_gene <- all_data_gene[, c(ncol(all_data_gene), 2:(ncol(all_data_gene) - 1))]

#filtering the dataset to keep only the highly expressed genes
filtered_data <- all_data_gene %>%
  rowwise() %>%
  filter(sum(c_across(starts_with("C"))) > 0) %>%
  filter(sum(c_across(starts_with("C"))) >= 30) %>%
  filter(mean(c_across(starts_with("C")) > 50) >= 0.5) %>%
  ungroup()  # Ungroup after rowwise operations

#removing the patients C547 and C548 from the dataset
filtered_data <- filtered_data %>%
  dplyr::select(-contains("C547"), -contains("C548"), -contains("C555"), -contains("C519"), -contains("C536"), -contains("C562"))

#extracting unique patient identifiers by taking the first part of the column names
patient_groups <- unique(gsub("_.*", "", colnames(filtered_data)[-1]))

#initialize a list to store the LOOCV data frames
loocv_dfs <- list()

#function to calculate the mean, variance, and CV for a given patient's data
calculate_stats <- function(patient_data) {
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(
      Mean = mean(c_across(where(is.numeric)), na.rm = TRUE),
      Variance = var(c_across(where(is.numeric)), na.rm = TRUE),
      CV = ifelse(Mean == 0, NA, sqrt(Variance) / Mean)  # Avoid division by zero
    ) %>%
    ungroup()  # Ungroup after calculations
  
  return(pat_stats)
}

#defining a list of patient-sample exclusions for LOOCV
exclusions <- list(
  C530 = "C530_B1_G1_CV",
  C538 = "C538_C1_G1_CV",
  C543 = "C543_B1_G10_CV",
  C551 = "C551_A1_G5_CV",
  C554 = "C554_C1_G3_CV",
  C560 = "C560_B1_G4_CV",
  C561 = "C561_E1_G1_CV",
  C518 = "C518_B1_G1_CV",
  C532 = "C532_D1_G2_CV",
  C528 = "C528_C1_G2_CV"
)
#loop through all patient groups and perform LOOCV where necessary
for (pat in patient_groups) {
  # Match the samples for each patient
  pat_data_filtered <- filtered_data %>%
    dplyr::select(hgnc_symbol, matches(paste0("^", pat, "_")))
  
  #getting the column names corresponding to the patient samples
  pat_columns <- colnames(pat_data_filtered)[grep(paste0("^", pat, "_"), colnames(pat_data_filtered))]
  
  #calculating the row-wise means (mean expression) across patient columns
  row_means <- rowMeans(pat_data_filtered[, -1], na.rm = TRUE)
  
  #calculating the median of the row-wise means
  median_threshold <- median(row_means, na.rm = TRUE)

  #checking if there's an exclusion for this patient
  if (!is.null(exclusions[[pat]])) {
    #excluding the specific sample for LOOCV (only once)
    loocv_data <- pat_data_filtered %>% ungroup() %>% dplyr::select(-contains(exclusions[[pat]]))
  } else {
    
    loocv_data <- pat_data_filtered
  }
  
  #checking if there are sufficient samples remaining to calculate CV
  if (ncol(loocv_data) <= 2) {
    message(paste("Skipping patient", pat, "due to insufficient samples for LOOCV."))
    next
  }
  
  #calculating Mean, Variance, and CV for the remaining samples
  stats <- calculate_stats(loocv_data)
  
  #filtering out genes with mean expression less than the median threshold
  stats_filtered <- stats %>%
    filter(Mean >= median_threshold)
  
  #selecting only the top 500 genes with the highest CV
  top_stats <- stats_filtered %>%
    arrange(desc(CV)) %>%
    slice_head(n = 500)

  if (nrow(top_stats) > 0) {
    cv_column_name <- paste(pat, "LOOCV_CV", sep = "_")
    loocv_dfs[[cv_column_name]] <- top_stats$CV
  }
  
  if (!is.null(loocv_dfs[[paste0(pat, "_LOOCV_CV")]])) {
    print(paste("LOOCV data frame for patient:", pat))
    print(top_stats) 
  }
}

```

2. FINDING THE RECURRENT GENES IN ALL THE DATAFRAMES. 

Note : These genes have been stored into the excel file called: "common_genes_epicc" and has been added to the Github.


```{r}
library(dplyr)
if (nrow(top_stats) > 0) {
  cv_column_name <- paste(pat, "LOOCV_CV", sep = "_")
  loocv_dfs[[cv_column_name]] <- top_stats %>% dplyr::select(hgnc_symbol, CV)
}

gene_lists <- lapply(loocv_dfs, function(df) {
  if (is.data.frame(df) && "hgnc_symbol" %in% colnames(df)) {
    return(df$hgnc_symbol)
  } else {
    return(NULL)
  }
})

gene_lists <- Filter(Negate(is.null), gene_lists)

#finding recurrent genes across all valid data frames
if (length(gene_lists) > 0) {
  recurrent_genes <- Reduce(intersect, gene_lists)
  print("Recurrent genes across all LOOCV data frames:")
  print(recurrent_genes)
} else {
  print("No valid data frames with 'hgnc_symbol' column found.")
}


```

3. Performing the pathway analysis for the common genes using enrichR. 

```{r}
# Install required packages
if (!requireNamespace("enrichR", quietly = TRUE)) {
  install.packages("enrichR")
}

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}

# Load libraries
library(enrichR)
library(ggplot2)

# Load your genes from the Excel file
library(readxl)

# Load the Excel file
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/common_genes_epicc.xlsx"  
genes_df <- read_excel(file_path)

# Assuming the gene names are in the first column
gene_list <- genes_df[[1]]  # Adjust the column index if necessary

# Perform enrichment analysis using Enrichr
dbs <- listEnrichrDbs()
db_to_use <- c("WikiPathways_2024_Human")

enrichr_results <- enrichr(gene_list, db_to_use)

# Extract the WikiPathways results for plotting
wiki_results <- enrichr_results[["WikiPathways_2024_Human"]]

# Sort the results by Adjusted P-value in ascending order
sorted_wiki <- wiki_results[order(wiki_results$Adjusted.P.value), ]

# Select the top pathways (e.g., top 10)
top_wiki <- sorted_wiki[1:10, ]

# Create a dot plot with Adjusted P-value on the x-axis
ggplot(top_wiki, aes(x = reorder(Term, -Adjusted.P.value), y = -log10(Adjusted.P.value))) +
  geom_point(aes(size = Combined.Score, color = Adjusted.P.value)) +
  coord_flip() +
  labs(x = "Pathway", y = "-log10(Adjusted P-value)", title = "Top 10 Enriched WikiPathways ") +
  theme_minimal()



```

4. Performing the gene ontology analysis and plotting all the three categories for the genes enriched in the first pathway. 

```{r}
library(enrichR)
library(dplyr)

# Define the gene list as a character vector
gene_list <- c("CASP7", "TRAF3", "TRAF2", "RAC1", "MAPK14", "MAP3K7", "BIRC3")

# Connect to Enrichr database and list available databases
dbs <- listEnrichrDbs()

# Select the GO terms databases (BP, CC, MF)
databases <- c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023")

# Perform enrichment analysis using the gene list for selected GO categories
enrichment_results <- enrichr(gene_list, databases)

# Extract and sort the GO Biological Process dataframe by ascending p-adjusted values
go_bp_df <- as.data.frame(enrichment_results[["GO_Biological_Process_2023"]]) %>%
  arrange(Adjusted.P.value)

# Extract and sort the GO Cellular Component dataframe by ascending p-adjusted values
go_cc_df <- as.data.frame(enrichment_results[["GO_Cellular_Component_2023"]]) %>%
  arrange(Adjusted.P.value)

# Extract and sort the GO Molecular Function dataframe by ascending p-adjusted values
go_mf_df <- as.data.frame(enrichment_results[["GO_Molecular_Function_2023"]]) %>%
  arrange(Adjusted.P.value)

# Print the sorted data frames
cat("GO Biological Process 2023 Results:\n")
print(go_bp_df)

cat("\nGO Cellular Component 2023 Results:\n")
print(go_cc_df)

cat("\nGO Molecular Function 2023 Results:\n")
print(go_mf_df)

```

```{r}
top10_bp <- go_bp_df %>% head(10)
top10_cc <- go_cc_df %>% head(10)
top10_mf <- go_mf_df %>% head(10)


# Plot for top 10 GO Biological Process with terms ordered by -log10(Adjusted.P.value)
bp_plot <- ggplot(top10_bp, aes(x = reorder(Term, -log10(Adjusted.P.value)), y = -log10(Adjusted.P.value), fill = "Biological Process")) +
  geom_bar(stat = "identity", fill = "#66c2a5") +  # Choose a color for BP
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 GO Biological Process",
       x = "GO Terms",
       y = "-log10(Adjusted P-value)") +
  scale_fill_manual(values = "#66c2a5")

# Save the plot for Biological Process
ggsave("C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/top10_GO_Biological_Process_1st_pathway.png", plot = bp_plot, width = 8, height = 6, bg = "white")

# Plot for top 10 GO Molecular Function with terms ordered by -log10(Adjusted.P.value)
mf_plot <- ggplot(top10_mf, aes(x = reorder(Term, -log10(Adjusted.P.value)), y = -log10(Adjusted.P.value), fill = "Molecular Function")) +
  geom_bar(stat = "identity", fill = "blue") +  
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 GO Molecular Function",
       x = "GO Terms",
       y = "-log10(Adjusted P-value)") +
  scale_fill_manual(values = "blue")

# Save the plot for Molecular Function
ggsave("C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/top10_GO_Molecular_Function_1st_pathway.png", plot = mf_plot, width = 8, height = 6, bg = "white")

# Plot for top 10 GO Cellular Component with terms ordered by -log10(Adjusted.P.value)
cc_plot <- ggplot(top10_cc, aes(x = reorder(Term, -log10(Adjusted.P.value)), y = -log10(Adjusted.P.value), fill = "Cellular Component")) +
  geom_bar(stat = "identity", fill = "#fc8d62") + 
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 GO Cellular Component",
       x = "GO Terms",
       y = "-log10(Adjusted P-value)") +
  scale_fill_manual(values = "#fc8d62")

# Save the plot for Cellular Component
ggsave("C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/top10_GO_Cellular_Component_1st_pathway.png", plot = cc_plot, width = 8, height = 6, bg = "white")



```


