---
title: "R Notebook"
output: html_notebook
---

Title : GENE ONTOLOGY ANALYSIS FOR THE IN HOUSE DATASET 


1. General Go Ontology analysis for the common genes - with respect to each ontology category ( Molecular Function , Biological Processes , and the Cellular Component). 
2. Performing the enrichment analysis using Enrichr. 

```{r}
#installing the package for enrichr database. 
if (!requireNamespace("enrichR", quietly = TRUE)) {
  install.packages("enrichR")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}

library(enrichR)
library(dplyr)
library(ggplot2)

#the common genes were stored 
# Load the dataset
input_file <- "C:/Users/Vaishnavi Raikar/Downloads/common_genes_crc.csv"
gene_list <- read.csv(input_file)

#converting the gene symbols into a vector
genes <- as.character(gene_list$Gene)

#chosing the GO databases to use
dbs <- c("GO_Biological_Process_2023", "GO_Molecular_Function_2023", "GO_Cellular_Component_2023")

# Perform the enrichment analysis for the GO terms 
enrichment_results <- enrichR::enrichr(genes, dbs)

go_bp_results <- enrichment_results[["GO_Biological_Process_2023"]]
go_mf_results <- enrichment_results[["GO_Molecular_Function_2023"]]
go_cc_results <- enrichment_results[["GO_Cellular_Component_2023"]]


#filetring all GO terms with Adjusted.P.value < 0.01 for each category

top_mf <- go_mf_results %>%
  filter(Adjusted.P.value < 0.01) %>%  
  arrange(Adjusted.P.value) %>%
  mutate(Category = "Biological Process")

# Filter and mutate for Molecular Function GO terms
top_bp <- go_bp_results %>%
  as.data.frame() %>%  # Ensure it's treated as a data frame
  filter(Adjusted.P.value < 0.01) %>%
  arrange(Adjusted.P.value) %>%
  head(20) %>%
  mutate(Category = "Molecular Function")

top_cc <- go_cc_results %>%
  filter(Adjusted.P.value < 0.01) %>%  
  arrange(Adjusted.P.value) %>%
  mutate(Category = "Cellular Component")

#visualizing GO Term and Pathway results

#Biological Process plot
bp_plot <- ggplot(top_bp, aes(x = reorder(Term, -log10(Adjusted.P.value)), y = -log10(Adjusted.P.value), fill = Category)) +
  geom_bar(stat = "identity", fill = "blue") +  
  coord_flip() +
  labs(title = "GO Terms vs Adj p-values - (BP)", 
       x = "GO Term", 
       y = "-log10(Adjusted P-value)") +
  theme_minimal() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.title = element_text(hjust = 0.5, size = 8))

#Molecular Function plot
mf_plot <- ggplot(top_mf, aes(x = reorder(Term, -log10(Adjusted.P.value)), y = -log10(Adjusted.P.value), fill = Category)) +
  geom_bar(stat = "identity", fill = "orange") +  
  coord_flip() +
  labs(title = "GO Terms vs Adj P-value (MF)", 
       x = "GO Term", 
       y = "-log10(Adj P-value)") +
  theme_minimal() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1),  
        axis.text.y = element_text(size = 8),               
        plot.title = element_text(hjust = 0.5, size = 10),  
        axis.title.x = element_text(size = 9),              
        axis.title.y = element_text(size = 9)) +            
  ylim(0, 6) 

#Cellular Component plot
cc_plot <- ggplot(top_cc, aes(x = reorder(Term, -log10(Adjusted.P.value)), y = -log10(Adjusted.P.value), fill = Category)) +
  geom_bar(stat = "identity", fill = "green") +  
  coord_flip() +
  labs(title = "GO Terms vs Adjusted p-values (CC)", 
       x = "GO Term", 
       y = "-log10(Adjusted P-value)") +
  theme_minimal() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.title = element_text(hjust = 0.5, size = 8))


print(cc_plot)
print(bp_plot)
print(mf_plot)
```

2. PERFORMING THE KEGG ENRICHMENT ANALYSIS FOR QUERYING THE TOP ENRICHED PATHWAYS FOR THE COMMON GENES. 

```{r}
#installing the necessary pacakages 
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("clusterProfiler")
BiocManager::install("org.Hs.eg.db")  #for human gene annotation
BiocManager::install("pathview")      #for visualizing pathways

library(clusterProfiler)
library(org.Hs.eg.db)
library(pathview)

#common genes excel file 
gene_data <- read.csv("C:/Users/Vaishnavi Raikar/Downloads/common_genes_crc.csv")

gene_list <- as.character(gene_data$Gene)

#converting the gene symbols to ENTREZ IDs (KEGG works with ENTREZ IDs)
#ensuring that the gene names are valid and map them to ENTREZ IDs
entrez_ids <- bitr(gene_list, fromType = "SYMBOL", 
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

#performing KEGG pathway enrichment analysis
kegg_enrich <- enrichKEGG(gene = entrez_ids$ENTREZID,
                          organism = 'hsa', # hsa refers to Homo sapiens
                          pvalueCutoff = 0.01)

#sorting the results by adjusted p-value (p.adjust) in ascending order
kegg_enrich@result <- kegg_enrich@result[order(kegg_enrich@result$p.adjust), ]


#plotting the KEGG pathway enrichment results 
dotplot(kegg_enrich, showCategory = 20)
```



3. FETCHING THE GENES ASSOCIATED WITH EACH OF THE TOP 3 ENRICHED PATHWAYS WITH LOWEST ADJUSTED.P.VALUE.
- All the genes are stored in the excel file called - "top_enriched_pathways_genes". 

```{r}
# Convert to dataframe
df <- data.frame(kegg_enrich)

#filtering for the row where ID is 'hsa04151'
genes_cyto <- df[df$ID == "hsa04820", ]
#filtering for the row where ID is 'hsa04151'
genes_pi3k <- df[df$ID == "hsa04151", ]
#filtering for the row where ID is 'hsa04151'
genes_focal <- df[df$ID == "hsa04510", ]


print(genes_cyto$geneID)

print(genes_pi3k$geneID)

print(genes_focal$geneID)

```
4. GO TERM ANALYSIS FOR THE GENES ASSOCIATED WITH EACH TOP 3 ENRICHED PATHWAYS 
1. The code needs to be changed for every pathway because of the genes associated with it, so on the 191st line , it just needs to be df[[1]] if you want to see the go terms for the Cytoskeleton pathway, df[[2]] if you want to see the go terms for the Pi3k signaling pathway , and df[[3]] for the focal adhesion pathway. 



```{r}

library(readxl)

#loading the dataset with the top enriched pathway genes 
input_file <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/top_enriched_pathways_genes.xlsx"
df <- read_xlsx(input_file)

#extracting the gene list from the first column (assuming it's named 'Gene' or similar)
gene_list <- as.character(df[[2]])  #1 :  Extract the first column as gene list
                                    #2 :  Extract the second column as gene list
                                    #3 :  Extract the third column as gene list

#connecting to Enrichr database and list available databases
dbs <- listEnrichrDbs()

#selecting the GO terms databases (BP, CC, MF)
databases <- c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023")

#performing enrichment analysis using the gene list for selected GO categories
enrichment_results <- enrichr(gene_list, databases)

#printing and storing each GO category in a separate data frame
go_bp_df <- as.data.frame(enrichment_results[["GO_Biological_Process_2023"]])
go_cc_df <- as.data.frame(enrichment_results[["GO_Cellular_Component_2023"]])
go_mf_df <- as.data.frame(enrichment_results[["GO_Molecular_Function_2023"]])


#sorting the GO Biological Process dataframe by ascending p-adjusted values
go_bp_df_sorted <- go_bp_df %>%
  arrange(Adjusted.P.value)

#sorting the GO Cellular Component dataframe by ascending p-adjusted values
go_cc_df_sorted <- go_cc_df %>%
  arrange(Adjusted.P.value)

#sorting the GO Molecular Function dataframe by ascending p-adjusted values
go_mf_df_sorted <- go_mf_df %>%
  arrange(Adjusted.P.value)

cat("GO Biological Process 2023 Results:\n")
print(go_bp_df_sorted)

cat("\nGO Cellular Component 2023 Results:\n")
print(go_cc_df_sorted)

cat("\nGO Molecular Function 2023 Results:\n")
print(go_mf_df_sorted)



```
5. PLOTTING THE GO TERMS BASED ON THE GENES ASSOCIATED WITH EACH PATHWAY. 

```{r}
#creating a bar plot for the top 10 GO terms based on ascending adjusted p-values
plot_go_terms <- function(go_df, title, color) {
  #selecting the top 10 terms based on Adjusted P-value (smallest values)
  top_go_df <- go_df[1:10,]
  
  #ensuring that the terms are ordered by ascending adjusted p-values (smallest p-value at top)
  top_go_df <- top_go_df %>%
    arrange(Adjusted.P.value)
  
  #Correctly order the factor levels for the terms to ensure smallest p-value is at the top
  top_go_df$Term <- factor(top_go_df$Term, levels = rev(top_go_df$Term))
  
  # then creating the bar plot with ascending adjusted p-values, placing the smallest on top
  ggplot(top_go_df, aes(x = Term, y = -log10(Adjusted.P.value))) +
    geom_bar(stat = "identity", fill = color) +
    coord_flip() +
    labs(title = title, x = "GO Term", y = "-log10(Adjusted P-value)") +
    theme_minimal(base_size = 10)
}

#plotting the bar plots for each category 
plot_bp <- plot_go_terms(go_bp_df_sorted, "Top 10 GO Biological Process Terms", "skyblue")

plot_cc <- plot_go_terms(go_cc_df_sorted, "Top 10 GO Cellular Component Terms", "lightgreen")

plot_mf <- plot_go_terms(go_mf_df_sorted, "Top 10 GO Molecular Function Terms", "salmon")

# Display the plots
print(plot_bp)
print(plot_cc)
print(plot_mf)


```

