------------------------------------------------------------------------

title: "R Notebook" output: html_notebook

Title : Performing the iCMS classification on the In-house dataset using the iCMS.SSC package from Github.

1.  INSTALLING THE PACKAGES FOR iCMS CLASSIFICATION.

```{r}
if(!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
library(devtools)

# Install the package from GitHub
devtools::install_github('CRCrepository/iCMS.SSC', force = TRUE)

```

2.  PERFORMING THE iCMS CLASSIFICATION ON THE DATASET AS HERE WE DO NOT NEED CONVERSION OF THE GENE SYMBOLS TO ENTREZ IDS.

```{r}
library(readxl)
library(writexl)
library(dplyr)
library(iCMS.SSC)

file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset.xlsx"

final_data <- read_excel(file_path)

#converting the data into a tibble
final_data_1 <- as_tibble(final_data)

#then to a dataframe to set row names
final_data_1 <- as.data.frame(final_data_1)

#checking if there are any non-numeric columns (excluding the gene column)
final_data_1[-1] <- lapply(final_data_1[-1], function(x) as.numeric(as.character(x)))

#setting the row names to gene symbols and remove the gene column
rownames(final_data_1) <- final_data_1$Gene
final_data_1 <- final_data_1 %>% dplyr::select(-Gene)

#matrix conversion
expression_matrix <- as.matrix(final_data_1)

#performing the iCMS classification on the entire matrix
icms_results <- iCMS.KNN(expression_matrix, jobs = 1)


print(icms_results)

```

3.  PLOTTING THE BAR PLOT FOR EACH PATIENT WITH THE COUNT OF SAMPLES BEING CLASSIFIED AS ANY OF THE TWO iCMS CATEGORIES.

```{r}

if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
library(readxl)

install.packages("ggplot2")
library("ggplot2")

if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
library(tidyr)

#uploading the iCMS_plot excel sheet where I stored the counts for each patient.
file_path <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/iCMS_plot.xlsx"

icms_results <- read_xlsx(file_path)

#plotting the data 
icms_long <- icms_results %>%
  pivot_longer(cols = c(starts_with("iCMS"), "NA"), names_to = "iCMS", values_to = "Count")

#defining the colors for the classification 
icms_colors <- c("iCMS2" = "#1f77b4", "iCMS3" = "yellow", "NA" = "grey" ) 

#plotting the bar plot
ggplot(icms_long, aes(x = Sample_Names, y = Count, fill = iCMS)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = icms_colors) +
  labs(title = "ICMS Counts per Patient Group",
       x = "Patient Group",
       y = "Count",
       fill = "CMS") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
