---
title: "R Notebook"
output: html_notebook
---
Title : CONTINUING THE EXPRESSION ANALYSIS FOR THE TOP 500 GENES WITH HIGHEST CV. 

1. PERFORMING LOOCV ( LEAVE OUT ONE COEFFICIENT OF VARIATION )

1. Expression analysis for the top 500 genes with the highest CV. 
2. Performing the LOOCV for each patient to understand if there is a striking variation due to any of the patient samples in each patient. 
3. For every iteration , one sample is being left out and the CV is calculated based on the rest of the patient samples. 

Note : the steps are repeated for filtering the data and the CV for a better understanding. The only changes are with respect to each blocks title. 

```{r}
library(dplyr)
library(readxl)

# Load the dataset
input_file <- "c:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - vaishnavi/In house dataset.xlsx"
input_data <- read_xlsx(input_file)

#removing the genes with 0 counts for the expression. 
filtered_data <- input_data %>%
  filter(rowSums(across(where(is.numeric))) > 0)

#filtering criteria for low expression genes or low variance genes
filtered_data <- filtered_data %>%
  filter_if(is.numeric, all_vars(. > 20))  #Filter for genes with expression > 20

#extracting the patient identifiers.
patient_groups <- unique(gsub("_Sam.*", "", colnames(filtered_data)[-1]))

#initializing a list to store the LOOCV data frames
loocv_dfs <- list()
average_cv_dfs <- data.frame()

#then calculate the mean, variance, and CV
calculate_stats <- function(patient_data) {
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(
      Mean = mean(c_across(where(is.numeric)), na.rm = TRUE),
      Variance = var(c_across(where(is.numeric)), na.rm = TRUE),
      CV = ifelse(Mean == 0, NA, sqrt(Variance) / Mean)  
    ) %>%
    ungroup()
  
  return(pat_stats)
}

#looping through each patient group to perform LOOCV
for (pat in patient_groups) {
  #matching the samples for each patient
  pat_data_filtered <- filtered_data %>%
    dplyr::select(Gene, matches(paste0("^", pat, "_Sam")))

  #getting the column names corresponding to the patient samples
  pat_columns <- colnames(pat_data_filtered)[grep(paste0("^", pat, "_Sam"), colnames(pat_data_filtered))]
  
  #initialising a list to store CV values for each LOOCV iteration
  cv_values <- list()
  
  #performing LOOCV for each sample in the patient group
  for (i in seq_along(pat_columns)) {
    # Exclude one sample at a time
    loocv_data <- pat_data_filtered %>% dplyr::select(-all_of(pat_columns[i]))
    
    #calculating the Mean, Variance, and CV for the remaining samples
    stats <- calculate_stats(loocv_data)
    
    #filtering out the genes with mean expression less than 100
    stats_filtered <- stats %>%
      filter(Mean >= 100)
    
    #selecting only the top 500 genes with the highest CV
    top_stats <- stats_filtered %>%
      arrange(desc(CV)) %>%
      slice_head(n = 500)
    
    #saving the CV values with the sample name left out in the label
    if (nrow(top_stats) > 0) {  # Ensure we have data before proceeding
      cv_column_name <- paste(pat_columns[i], "CV", sep = "_")
      loocv_dfs[[cv_column_name]] <- top_stats$CV
      cv_values[[cv_column_name]] <- top_stats$CV
    }
  }
  
  #combining the CV values into a data frame if there are CV values
  if (length(cv_values) > 0) {
    cv_df <- as.data.frame(cv_values)
    
    #calculating the average CV for each LOOCV
    avg_cv <- colMeans(cv_df, na.rm = TRUE)
   
    if (length(avg_cv) > 0) {
      #combing the average CV into the result data frame
      avg_cv_df <- data.frame(Patient = pat, Sample = names(avg_cv), Avg_CV = avg_cv)
      
      average_cv_dfs <- rbind(average_cv_dfs, avg_cv_df)
    }
  }
}

```


2. PLOTTING THE LOOCV PLOT AND AVERAGE CV AFTER LOOCV. 

```{r}
install.packages("RColorBrewer")
library(RColorBrewer)
library(ggplot2)

average_cv_dfs$Sample <- factor(average_cv_dfs$Sample, levels = unique(average_cv_dfs$Sample))

#defining the set of color palette for the LOOCV plot
my_colors <- brewer.pal(5, "Dark2")

#plotting the samples
ggplot(average_cv_dfs, aes(x = Patient, y = Avg_CV, color = Sample)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Average CV per Patient", 
       x = "Patient", 
       y = "Average CV") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = rep(my_colors, length.out = length(unique(average_cv_dfs$Sample))))


#creating a new data frame to store overall averages for each patient
overall_avg_cv <- average_cv_dfs %>%
  group_by(Patient) %>%
  summarise(Overall_Avg_CV = mean(Avg_CV, na.rm = TRUE))

#using a larger palette 
ggplot(overall_avg_cv, aes(x = Patient, y = Overall_Avg_CV, fill = Patient)) +
  geom_bar(stat = "identity", width = 0.6) +
  theme_minimal() +
  labs(title = "Overall Average CV for Each Patient - after LOOCV", 
       x = "Patient", 
       y = "Overall Average CV") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")  # Using a larger color palette



```


3. ONLY PERFORMING LOOCV FOR PAT_1 AND PAT_3 BECAUSE THESE PATIENTS SHOWED CLEAR OUTLIERS AND FOR THE REST OF THE PATIENTS THE CV IS CALCULATED OVERALL. NEXT, THE DATAFRAMES ARE PRINTED WITH PAT_1 ONLY WITH 4 SAMPLES BY REMOVING THE OLUTLIER PATIENT SAMPLE. SIMILARILY FOR PAT_3. 

```{r}
library(dplyr)
library(readxl)

# Loading the dataset
input_file <- "c:/Users/Vaishnavi Raikar/OneDrive/Desktop/Major Thesis - Vaishnavi/In house dataset.xlsx"
input_data <- read_xlsx(input_file)

numeric_columns <- input_data %>%
  select(where(is.numeric))  

#rowSums to only numeric columns
filtered_data <- input_data %>%
  filter(rowSums(numeric_columns) > 0)


filtered_data <- filtered_data %>%
  filter_if(is.numeric, all_vars(. > 20))

patient_groups <- unique(gsub("_Sam.*", "", colnames(filtered_data)[-1]))

#Initializing the list to store the LOOCV data frames
loocv_dfs <- list()

#then calculate the mean, variance, and CV
calculate_stats <- function(patient_data) {
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(
      Mean = mean(c_across(where(is.numeric)), na.rm = TRUE),
      Variance = var(c_across(where(is.numeric)), na.rm = TRUE),
      CV = ifelse(Mean == 0, NA, sqrt(Variance) / Mean)  # Avoid division by zero
    ) %>%
    ungroup()
  
  return(pat_stats)
}

#looping through each patient group
for (pat in patient_groups) {
 
  pat_data_filtered <- filtered_data[, c(1, grep(paste0("^", pat, "_Sam"), colnames(filtered_data)))]
  
  if (pat %in% c("Pat1", "Pat3")) {
    # only perform LOOCV for Pat1 and Pat3
    pat_columns <- colnames(pat_data_filtered)[grep(paste0("^", pat, "_Sam"), colnames(pat_data_filtered))]
    cv_values <- list()
    
    for (i in seq_along(pat_columns)) {
      # Use base R to exclude one sample at a time
      loocv_data <- pat_data_filtered[, -which(colnames(pat_data_filtered) == pat_columns[i])]
      
      #checking if loocv_data has any rows left before proceeding
      if (ncol(loocv_data) > 1) {
        stats <- calculate_stats(loocv_data)
        stats_filtered <- stats %>% filter(Mean >= 100)
      
        if (nrow(stats_filtered) > 0) {
          top_stats <- stats_filtered %>%
            arrange(desc(CV)) %>%
            slice_head(n = 500)
          
          #saving the CV values with proper name
          cv_column_name <- paste(pat, "CV", i, sep = "_")
          loocv_dfs[[cv_column_name]] <- top_stats
        }
      }
    }
    
  } else {
    # Calculating the mean, variance, CV for the rest of the patients
    if (ncol(pat_data_filtered) > 1) {
      stats <- calculate_stats(pat_data_filtered)
      stats_filtered <- stats %>% filter(Mean >= 100)
      
      #Ensuring there are filtered results before proceeeding
      if (nrow(stats_filtered) > 0) {
        top_stats <- stats_filtered %>%
          arrange(desc(CV)) %>%
          slice_head(n = 500)
        
        #saving the top 500 genes with the highest CV
        loocv_dfs[[paste(pat, "Overall", sep = "_")]] <- top_stats
      }
    }
  }
}

#printing the dataframes
selected_samples <- c("Pat1_CV_4", "Pat3_CV_1", "Pat2_Overall", "Pat4_Overall", 
                      "Pat5_Overall", "Pat6_Overall", "Pat7_Overall", 
                      "Pat8_Overall", "Pat9_Overall", "Pat10_Overall")

for (sample in selected_samples) {
  cat(paste("Data Frame for", sample, ":\n"))
  
  if (!is.null(loocv_dfs[[sample]])) {
    print(loocv_dfs[[sample]])
  } else {
    cat("No data available for this sample.\n")
  }
  cat("\n")  
}



```


4. IDENTIFYING THE RECURRENT GENES IN ALL THESE PATIENTS. SO IF THE GENE IS OCCURING IN MORE THAN 3 PATIENTS , THE GENE IS PRINTED. 


```{r}
# 1.combining the gene lists from selected samples into a single data frame
all_genes <- bind_rows(lapply(selected_samples, function(sample) {
  data.frame(Gene = loocv_dfs[[sample]]$Gene, Patient = sample)
}))

# 2.identifying the common genes that appear in at least 3 different patient samples
common_genes <- all_genes %>%
  group_by(Gene) %>%
  summarize(Count = n_distinct(Patient)) %>%
  filter(Count >= 3) %>%
  arrange(desc(Count))

# 3.storing the common genes by initialising a new dataframe 
common_genes_cvs <- data.frame(Gene = common_genes$Gene)

# 4.looping through selected_samples and add the CV values for each common gene
for (sample in selected_samples) {
  # Extract the Gene and CV columns for the current sample
  sample_df <- loocv_dfs[[sample]][, c("Gene", "CV")]
  
  #merging current sample's CV values into the common_genes_cvs dataframe
  common_genes_cvs <- merge(common_genes_cvs, sample_df, by = "Gene", all.x = TRUE)
  
  #rennaming the CV column to correspond to the sample name
  colnames(common_genes_cvs)[ncol(common_genes_cvs)] <- sample
}

# 5.replacing the NA values with 0 
common_genes_cvs[is.na(common_genes_cvs)] <- 0
print(common_genes_cvs)

# Assuming final_data is your DataFrame
write.csv(final_data, "C:/Users/Vaishnavi Raikar/Downloads/common_genes_crc.csv", row.names = FALSE)

# Load necessary libraries
library(ggplot2)

```

5. ( Additional analysis code ) : PLOTTING A HEATMAP TO CHECK THE RECURRENT GENES BETWEEN PAIR OF PATIENTS. 


```{r}
library(dplyr)
library(readxl)
library(tidyr)  

#loading the dataset
input_file <- "C:/Users/Vaishnavi Raikar/OneDrive/Desktop/MAJOR THESIS - VAISHNAVI/In house dataset.xlsx"
input_data <- read_xlsx(input_file)

#filtering data to remove genes with 0 counts across all samples
filtered_data <- input_data %>%
  filter(rowSums(select(., where(is.numeric))) > 0)

#filtering out genes with counts less than 20 in any patient sample
filtered_data <- filtered_data %>%
  filter_if(is.numeric, all_vars(. > 20))

#unique patient identifiers
patient_groups <- unique(gsub("_Sam.*", "", colnames(filtered_data)[-1]))

#store the LOOCV data frames in a list
loocv_dfs <- list()

calculate_stats <- function(patient_data) {
  #select only numeric columns for the calculations
  numeric_patient_data <- patient_data %>% 
    select(where(is.numeric))
  
  #calculate Mean, Variance, and CV
  pat_stats <- patient_data %>%
    rowwise() %>%
    mutate(
      Mean = ifelse(any(!is.na(c_across(all_of(colnames(numeric_patient_data))))),
                    mean(c_across(all_of(colnames(numeric_patient_data))), na.rm = TRUE),
                    NA),
      Variance = ifelse(any(!is.na(c_across(all_of(colnames(numeric_patient_data))))),
                        var(c_across(all_of(colnames(numeric_patient_data))), na.rm = TRUE),
                        NA),
      CV = ifelse(Mean == 0 | is.na(Mean), NA, sqrt(Variance) / Mean)
    ) %>%
    ungroup()
  
  return(pat_stats)
}

#loop through each patient group
for (pat in patient_groups) {
  pat_data_filtered <- filtered_data %>%
    select(Gene, matches(paste0("^", pat, "_Sam")))
  
  if (pat %in% c("Pat1", "Pat3")) {
    pat_columns <- colnames(pat_data_filtered)[grep(paste0("^", pat, "_Sam"), colnames(pat_data_filtered))]
    
    for (i in seq_along(pat_columns)) {
      loocv_data <- pat_data_filtered %>% select(-all_of(pat_columns[i]))
      stats <- calculate_stats(loocv_data)
      stats_filtered <- stats %>% filter(Mean >= 100)
      top_stats <- stats_filtered %>% arrange(desc(CV)) %>% slice_head(n = 500)
      cv_column_name <- paste(pat, "CV", i, sep = "_")
      loocv_dfs[[cv_column_name]] <- top_stats
    }
    
  } else {
    stats <- calculate_stats(pat_data_filtered)
    stats_filtered <- stats %>% filter(Mean >= 100)
    top_stats <- stats_filtered %>% arrange(desc(CV)) %>% slice_head(n = 500)
    loocv_dfs[[paste(pat, "Overall", sep = "_")]] <- top_stats
  }
}

#select the specific samples for further analysis
selected_samples <- c("Pat1_CV_4", "Pat3_CV_1", "Pat2_Overall", "Pat4_Overall", 
                      "Pat5_Overall", "Pat6_Overall", "Pat7_Overall", 
                      "Pat8_Overall", "Pat9_Overall", "Pat10_Overall")

#extracting  the gene lists
gene_lists <- lapply(selected_samples, function(sample) loocv_dfs[[sample]]$Gene)

#combining all the gene lists into a single data frame with patient information
all_genes <- bind_rows(lapply(selected_samples, function(sample) {
  data.frame(Gene = loocv_dfs[[sample]]$Gene, Patient = sample)
}))

#finding genes that are present in more than one patient
recurrent_genes <- all_genes %>%
  group_by(Gene) %>%
  summarize(Count = n_distinct(Patient)) %>%
  filter(Count > 1) %>%
  arrange(desc(Count))

#preparing the data for the heatmap
heatmap_data <- all_genes %>%
  filter(Gene %in% recurrent_genes$Gene) %>%
  mutate(Present = 1) %>%
  pivot_wider(names_from = Patient, values_from = Present, values_fill = 0)

#converting the data frame to a matrix for the heatmap
gene_matrix <- as.matrix(heatmap_data[,-1])  # Remove the Gene column

#setting the row names to the Gene names
rownames(gene_matrix) <- heatmap_data$Gene


install.packages("pheatmap")
library(pheatmap)
#preparing the data for the heatmap by finding intersections between each pair of patients
patient_pairs <- combn(selected_samples, 2, simplify = FALSE)

#Initializing a matrix to store the number of shared genes between patients
shared_genes_matrix <- matrix(0, nrow = length(selected_samples), ncol = length(selected_samples))
rownames(shared_genes_matrix) <- selected_samples
colnames(shared_genes_matrix) <- selected_samples

#filling the matrix with the number of shared genes
for (pair in patient_pairs) {
  patient1_genes <- loocv_dfs[[pair[1]]]$Gene
  patient2_genes <- loocv_dfs[[pair[2]]]$Gene
  shared_genes <- length(intersect(patient1_genes, patient2_genes))
  
  shared_genes_matrix[pair[1], pair[2]] <- shared_genes
  shared_genes_matrix[pair[2], pair[1]] <- shared_genes  # Symmetric matrix
}

diag(shared_genes_matrix) <- 500

#defining a color palette where 500 is white, and the rest follows a smooth gradient
color_palette <- c("white", colorRampPalette(c("lavender", "violet", "white"))(49))


breaks_list <- c(seq(0, 499, length.out = 50), 500)

#creating the heatmap with the adjusted gradient
pheatmap(shared_genes_matrix,
         cluster_rows = FALSE,   
         cluster_cols = FALSE,   
         color = color_palette,  
         breaks = breaks_list,   
         display_numbers = TRUE, 
         main = "Intersection of Top-CV Genes Between Patients")


```

6. PLOTTING A HISTOGRAM TO SEE HOW MANY TIMES EACH GENE APPEARS ACROSS THE SELECTED PATIENTS. 

```{r}
#extracting the gene columns from each of the selected samples
gene_lists <- lapply(selected_samples, function(sample) loocv_dfs[[sample]]$Gene)

#combining the genes into a single data frame with a column indicating the patient
all_genes <- bind_rows(lapply(selected_samples, function(sample) {
  data.frame(Gene = loocv_dfs[[sample]]$Gene, Patient = sample)
}))

#counting how many times each gene appears across the selected patients
gene_recurrence <- all_genes %>%
  group_by(Gene) %>%
  summarize(Recurrence = n_distinct(Patient))
#lastly plotting histogram with correct binwidth and integer x-axis
ggplot(gene_recurrence, aes(x = Recurrence)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1, max(gene_recurrence$Recurrence), by = 1)) +  # Ensure whole numbers on x-axis
  labs(title = "Histogram of Gene Recurrence Across Patients",
       x = "Number of Patients",
       y = "Number of Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```




